Late and early effective start of non-pharmaceutical interventions inferred during COVID-19 outbreaks

### 2020-05-11 ms/ms.tex
delayed-advanced -> late-early

Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
We compare these estimates to the official dates and find both delayed and advanced effect of NPIs on COVID\=/19 transmission dynamics.
We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework. % TODO explain why important

@@

In the following, we describe our findings on delayed and advanced start of NPI in detail.

%%% Delayed
\paragraph*{Delayed effective start of NPI.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).

@@

%%% Fig 2 %%%
\begin{figure}[h]
    \centering
    % Wuhan
    % τ_hat = 19.33
	% +-2.65 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
    \end{subfigure}
  	~
	% Italy
	% τ_hat = 24.74
	% +-0.71 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
    \end{subfigure}
    \caption{
	\textbf{Delayed effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:fig2}
\end{figure}

%%% Advanced
\paragraph*{Advanced effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).

@@

%%% Fig 3 %%%
\begin{figure}[h]
    \centering
    % France
    % τ_hat = 12.38
	% +-5.90 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3a.pdf}
    \end{subfigure}
    ~
    % Spain
    % τ_hat = 13.95
	% +-2.96 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3b.pdf}
    \end{subfigure}
    \caption{
    \textbf{Advanced effect of non-pharmaceutical interventions in France and Spain.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:fig3}
\end{figure}

@@

%%% Assessment
\paragraph*{Effect of delays and advances of real-time assessment.}

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the delayed or advanced population response to the NPI.
Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its delayed effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line. However, the number of cases will actually follow the black line, leading to a significant different ($\Delta$) between the projections and the realization.

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Delayed effective start of NPI leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). Model predictions, assuming a 50\% decrease in transmission rate after the NPI starts, are shown as colored lines with 95\% confidence intervals. 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are MAP estimates for Italy (\hl{TABLE}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
    } 
    \label{fig:fig4}
\end{figure}

@@

We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
We find examples of both advanced and delayed response to NPIs (\autoref{fig:fig1}).

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred 3-5 after the official date (\autoref{fig:fig2}). This could be explained by low compliance. In Italy, for example, a leak about the intent to lockdown Northern provinces results in people leaving those provinces~\citep{Gatto2020}. However, delayed effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.

In contrast, in most investigated countries, such as Spain and France, transmission rates seem to have been reduced even before official lockdowns were implemented (\autoref{fig:fig3}).
This advanced response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that milder measures that were adopted by the population, possibly due to government recommendations, media coverage, and social networks, could have been sufficient for epidemic control.
\hl{check if this is true} Indeed, the evidence supports a change in transmission dynamics (i.e. a model with $\tau$) even for Sweden, in which a lockdown was not implemented, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020} (Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.)

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
However, the delays and advances we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).

>>>>

Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
We compare these estimates to the official dates and find both late and early effects of NPIs on COVID\=/19 transmission dynamics.
We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework. % TODO explain why important

@@

In the following, we describe our findings on late and early effective start of NPI in detail.

%%% Late
\paragraph*{Late effective start of NPIs.}
In both Wuhan, China, and in Italy we find that our estimated effective start of NPI $\hat{\tau}$ is significantly later than the official date $\tau^*$ (\autoref{fig:fig1}).

@@

%%% Fig 2 %%%
\begin{figure}[h]
    \centering
    % Wuhan
    % τ_hat = 19.33
	% +-2.65 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
    \end{subfigure}
  	~
	% Italy
	% τ_hat = 24.74
	% +-0.71 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:fig2}
\end{figure}

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).

@@

%%% Fig 3 %%%
\begin{figure}[h]
    \centering
    % France
    % τ_hat = 12.38
	% +-5.90 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3a.pdf}
    \end{subfigure}
    ~
    % Spain
    % τ_hat = 13.95
	% +-2.96 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig3b.pdf}
    \end{subfigure}
    \caption{
    \textbf{Early effect of non-pharmaceutical interventions in France and Spain.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
	}
	\label{fig:fig3}
\end{figure}

@@

%%% Assessment
\paragraph*{Effect of late and early effect of NPIs on real-time assessment.}

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
Such a case is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$, but its late effect on the transmission dynamics starts on Mar~15 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases to follow the dashed red line. However, the number of cases will actually follow the black line, leading to a significant different ($\Delta$) between the projections and the realization.

%%% Fig 4 %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig4.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). Model predictions, assuming a 50\% decrease in transmission rate after the NPI starts, are shown as colored lines with 95\% confidence intervals. 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are MAP estimates for Italy (\hl{TABLE}) but with $\lambda=0.5$ and $\alpha_1=\alpha_2$.
    } 
    \label{fig:fig4}
\end{figure}

@@

We have estimated the effective start date of NPIs in several geographical regions using an SEIR epidemiological model and an MCMC parameter estimation framework.
We find examples of both late and early effect of NPIs (\autoref{fig:fig1}).

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred 3-5 after the official date (\autoref{fig:fig2}). This could be explained by low compliance. In Italy, for example, a leak about the intent to lockdown Northern provinces results in people leaving those provinces~\citep{Gatto2020}. However, late effect of NPIs could also be due to the time required by both the government and the citizens to organize for a lockdown.

In contrast, in most investigated countries, such as Spain and France, transmission rates seem to have been reduced even before official lockdowns were implemented (\autoref{fig:fig3}).
This early response is possibly due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response increased risk perception due to domestic or international COVID\=/19-related reports.
This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that milder measures that were adopted by the population, possibly due to government recommendations, media coverage, and social networks, could have been sufficient for epidemic control.
\hl{check if this is true} Indeed, the evidence supports a change in transmission dynamics (i.e. a model with $\tau$) even for Sweden, in which a lockdown was not implemented, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020} (Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.)

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a 7 day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of COVID\=/19~\citep{Gatto2020}.
However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).
